---
title: "Cyclistic Bike-Share Case Study"
author: "Olivia Chen"
date: "2023-09-08"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Cyclistic is a fictional company, which is a bike-share company in Chicago. It launched a bike-share program with flexibility of pricing plans for casual riders and annual members. In this study, we are trying to understand the differences between casual riders and annual members to help the team to design a new marketing strategy to maximize the number of annual memberships.


## Data
The Cyclistic's monthly trip data were obtained from an open source created by Motivate International Inc. The combined data set contains 3,158,109 records between January 2023 and July 2023 and 13 variables. 

### Set-up

```{r, message = FALSE, warning= FALSE}
library(tidyverse)
library(skimr)
library(dplyr)
library(ggmap)
library(ggplot2)
```

### Combine dataset

```{r pressure, message = FALSE, warning= FALSE}
setwd("C:/Users/Olivia/Desktop/MSBA 21/R/case study/")
tripdata <- 
  list.files(path = "C:/Users/Olivia/Desktop/MSBA 21/R/case study/", pattern = "*.csv") %>% 
  map_df(~read_csv(.))
```

### Data Understanding and Preprocessing
```{r warning= FALSE}
summary(tripdata)
#check null values
colSums(is.na(tripdata))
#calculate ops hours
tripdata$ride_length <- with(tripdata, 
                                    difftime(tripdata$ended_at, tripdata$started_at,
                                             units = "hours"))
tripdata$ride_length <- as.numeric(tripdata$ride_length, unites='hours')

#get date of the week (1=Sunday, 7=Saturday)
tripdata$day_of_week <- wday(tripdata$started_at)
#create subset of the data
sub_tripdata <- select(tripdata,
                                c('ride_id','rideable_type','member_casual','ride_length','day_of_week'))

#outliers
ggplot(sub_tripdata) + aes(x='',y=ride_length) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()

#remove inssuficient ride_length
sub_tripdata <- subset(sub_tripdata,ride_length > 0)
```

### Calculate the mean of ride_length and the max ride_length
```{r}
summary(sub_tripdata$ride_length)
```
The average length of ride is about 18.5 mins (0.308 hours) and the maximum length of ride is about 14.3 days (857.69 hours).

### Calculate the mode of day of week
```{r}
cal_mode <- function(x){which.max(tabulate(x))}
cal_mode(sub_tripdata$day_of_week)
```
Riders are more likely to utilize the bike-share service on Saturday as number 7 appears the most often in the data set.


```{r}
ggplot(data = sub_tripdata) +
  geom_bar(mapping = aes(x = member_casual, fill = rideable_type)) +
  labs(title="Cyclistic Bike-share: Member Casual vs Riderble Type")
```

The above bar chart indicates annual members has a stronger demand on both classic bike and electric bike comparing to casual riders.

```{r}
count(sub_tripdata,member_casual)
```
The above numbers indicate there is about 63.29% of Cyclistic users are annual members between January 2023 and July 2023. 

```{r}
ggplot(data=sub_tripdata) + geom_smooth(mapping=aes(x=ride_length, y=day_of_week), method = "gam")+
  geom_point(mapping=aes(x=ride_length, y=day_of_week,shape=member_casual, color = member_casual )) +
  labs(title="Cyclistic Bike-share: Ride Length vs Day of Week")
```

According to the above, casual riders, which are represented by red circle, tend to ride for a long period of time across all days of the week. While annual members, which are represented by blue triangles, concentrate in a relatively shorter ride length.  

### Create subset of the data for annual members
```{r}
annual_members <- filter(sub_tripdata, member_casual=="member")
summary(annual_members$ride_length)
```
The average ride length for annual members is about 12.4 minutes. 

```{r}
ggplot(data=annual_members) + geom_smooth(mapping=aes(x=ride_length, y=day_of_week), method = "gam")+
  geom_point(mapping=aes(x=ride_length, y=day_of_week,shape=member_casual )) +
  labs(title="Cyclistic Bike-share: Ride Length vs Day of Week",
       subtitle = "Annual Members Only")
```

Now, let's focus on the distributions of annual members. According to the chart above, the distributions of ride length of annual members are slightly right-skewed among all day of week. 


## Recommendations
Due to the higher demand on both classic bikes and electric bikes for annual members, I would recommend Cyclistic to reserve a few classic bikes and electric bikes in each station for annual members only. 
Since casual rider tend to ride for longer period of time, I would recommend to increase the price for single-ride passes which will encourage riders who use share-bike over 24 hours to convert to annual members.



## Next Steps
1) Utilize the location variables such as start_lat, start_lng, end_lat, end_lng to identify some target locations for marketing campaign.
2) Gather user information and consider launch incentive programs to maintain customers' loyalty. 
3) Since the average ride length for annual member is about 12.4 minutes, Cyclistic should consider investigate on the docking stations to ensure there will be enough docking stations near about 12 minutes radius from the popular/high-demand stations.
